\section*{Contributors}
\vspace{0.5em}
\begin{small}
\setlength{\LTleft}{0pt}
\setlength{\LTright}{0pt}
\renewcommand{\arraystretch}{1.3}
\newlength{\contribwidth}
\setlength{\contribwidth}{\dimexpr\textwidth-5.5cm-4\tabcolsep\relax}
\begin{longtable}{|l|l|}
\hline
Name & Contribution \\
\hline
\endfirsthead
\hline
Name & Contribution \\
\hline
\endhead
\hline
\endfoot
Camille Avestruz & \parbox[t]{\contribwidth}{Wrote a bulk of the original text of section 3.4 and contributed to section 3.8.} \\
Matthew R. Becker & \parbox[t]{\contribwidth}{Originated the idea to write the white paper and developed initial scope of content.} \\
Biswajit Biswas & \parbox[t]{\contribwidth}{Contributed to sec 3.8} \\
Rahul Biswas & \parbox[t]{\contribwidth}{Wrote text for section in 3.1 and contributed to the paper in general by suggesting edits.} \\
Adam S. Bolton & \parbox[t]{\contribwidth}{Coordinated and wrote most of the original text for the Infrastructure section} \\
Clecio R. Bom & \parbox[t]{\contribwidth}{Coordinated agentic AI subsection and contributed text along the document.} \\
Rapha\"el Bonnet-Guerrini & \parbox[t]{\contribwidth}{Contributed to text and editing section 4.1.} \\
Jean-Eric Campagne & \parbox[t]{\contribwidth}{Contributed text to sections 3.1, 3.2, 4.1, 4.2, 5.1} \\
Aleksandra \'Ciprijanovi\'c & \parbox[t]{\contribwidth}{Initial discussion and text review, contributed to sections 3 and 4} \\
Michael W. Coughlin & \parbox[t]{\contribwidth}{Contributed to early text of Section 6} \\
Juan C. Cuevas-Tello & \parbox[t]{\contribwidth}{Contributed to section 3.2} \\
Juan de Vicente & \parbox[t]{\contribwidth}{Contributed text to Section 3.1} \\
Steven Dillmann & \parbox[t]{\contribwidth}{Contributed text to and editing of Sections 5.1 and 5.2} \\
Alex Drlica-Wagner & \parbox[t]{\contribwidth}{Contributed to initial ideation and editing}\\
Sydney Erickson & \parbox[t]{\contribwidth}{Contributed text to and editing of Section 3.2} \\
Alexander T. Gagliano & \parbox[t]{\contribwidth}{Co-led the paper, contributed text and editing throughout the document.} \\
Christos Georgiou & \parbox[t]{\contribwidth}{CG contributed to the Theory and Modelling (3.6) and Introduction (2) section.} \\
Aritra Ghosh & \parbox[t]{\contribwidth}{Contributed to text in Sections 5 and 6} \\
Matthew Grayling & \parbox[t]{\contribwidth}{Contributed to text in Section 3.5} \\
Kirill A. Grishin & \parbox[t]{\contribwidth}{Contributed to the description of the use of variational auto-encoders in Section 3}\\
Mustapha Ishak & \parbox[t]{\contribwidth}{Contributed text to Section 3.6} \\
Wassim Kabalan & \parbox[t]{\contribwidth}{Contributed to sections 3.7 and 4.1.1} \\
Fran\c{c}ois Lanusse & \parbox[t]{\contribwidth}{Co-led the paper, contributed text and editing throughout the document.} \\
C. Danielle Leonard & \parbox[t]{\contribwidth}{Contributed to Section 3.6 (Theory and Modelling; Fast Surrogates for Cosmological Likelihoods).} \\
Michelle Lochner & \parbox[t]{\contribwidth}{Contributed significantly to the executive summary and conclusions, led section 5.1 (foundation models), contributing most of the text and editing of that section, wrote the text for section 4.3 (novelty detection and discovery), and contributed text, editing and ideas throughout the document. LSST DESC builder.} \\
Yao-Yuan Mao & \parbox[t]{\contribwidth}{Contributed to the initial ideation and Section 8} \\
Peter Melchior & \parbox[t]{\contribwidth}{Co-led section 4.} \\
Grant Merz & \parbox[t]{\contribwidth}{Contributed text to Section 3.8} \\
Martin Millon & \parbox[t]{\contribwidth}{Contributed text and editing to Section 3.2} \\
Anais M\"oller & \parbox[t]{\contribwidth}{Co-led section 4 and contributed text in Section 3.} \\
Gautham Narayan & \parbox[t]{\contribwidth}{Contributed text and editing throughout the document.} \\
Yuuki Omori & \parbox[t]{\contribwidth}{Contributed text to section 3.3}\\
Hiranya Peiris & \parbox[t]{\contribwidth}{Contributed text to sections 3.1, 3.3, 3.7, 4.1, and 4.2, and editing throughout the document.} \\
Laurence Perreault-Levasseur & \parbox[t]{\contribwidth}{Contributed text to section 3.2}\\
Andrés A. Plazas Malagón & \parbox[t]{\contribwidth}{Contributed text on Instrumental Response modeling. Rubin and LSST DESC builder.}\\
Nesar Ramachandra & \parbox[t]{\contribwidth}{Contributed text for Section 3.6 (Theory and Modelling)} \\
Benjamin Remy & \parbox[t]{\contribwidth}{Contributed text to sections 3.3 and 4.2} \\
Stefan Schuldt & \parbox[t]{\contribwidth}{Led section 3.2 (strong lensing), wrote most of its text, and provided input to other sections as well.\strut} \\
Ignacio Sevilla-Noarbe & \parbox[t]{\contribwidth}{Contributed text to sections 3, 7 and 8.} \\
Ved G. Shah & \parbox[t]{\contribwidth}{Contributed text to section 3, and editing throughout Section 3.} \\
Stephen Thorp & \parbox[t]{\contribwidth}{Contributed text to sections 3.1, 3.2, 3.3, 3.7, 4.1, 4.2 and 7, and editing throughout the document.} \\
Laura Toribio San Cipriano & \parbox[t]{\contribwidth}{Contributed text to section 3.} \\
Tilman Tr\"oster & \parbox[t]{\contribwidth}{Contributed to early drafts of sections 3.3 and 3.6} \\
Roberto Trotta & \parbox[t]{\contribwidth}{Contributed significant text and structure of section 4; some text to section 3; comments and suggestions throughout the paper} \\
Padma Venkatraman & \parbox[t]{\contribwidth}{Contributed text and edits to section 3.2} \\
Amanda Wasserman & \parbox[t]{\contribwidth}{Contributed text to section 3.5} \\
Tim White & \parbox[t]{\contribwidth}{Contributed to section 3.9} \\
Justine Zeghal & \parbox[t]{\contribwidth}{Contributed text to section 3.3 and 4.1.2} \\
Tianqing Zhang & \parbox[t]{\contribwidth}{Contributed to Section 3.1, and organized 3.8 and 3.9.} \\
Yuanyuan Zhang & \parbox[t]{\contribwidth}{Contributed text to Section 3.} \\
\end{longtable}
\end{small}
