% genera;
The DESC acknowledges ongoing support from the Institut National de Physique Nucl\'eaire et de Physique des Particules (IN2P3) in France; the Science \& Technology Facilities Council (STFC) in the United Kingdom; and the Department of Energy (DOE), the National Science Foundation (NSF), and the LSST Discovery Alliance (LSST-DA) in the United States. DESC uses resources of the IN2P3 Computing Center (CC-IN2P3--Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique; the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S.\ Department of Energy under Contract No.\ DE-AC02-05CH11231; STFC DiRAC HPC Facilities, funded by UK BEIS National E-infrastructure capital grants; and the UK particle physics grid, supported by the GridPP Collaboration. This work was performed in part under DOE Contract DE-AC02-76SF00515.
% authors' acknowledgments
A.T.G.\ is supported by the National Science Foundation under Cooperative Agreement PHY-2019786 (The NSF AI Institute for Artificial Intelligence and Fundamental Interactions, \url{http://iaifi.org/}). 
A.M. is supported by the Australian Research Council Discovery Early Career Research Award (DE230100055). 
M.L.\ acknowledges support from the South African Radio Astronomy Observatory and the National Research Foundation (NRF) towards this research. Opinions expressed and conclusions arrived at, are those of the authors and are not necessarily to be attributed to the NRF. 
H.P.\ and S.T.\ have been supported by funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programmes (grant agreement no.\ 101018897 CosmicExplorer). H.P.\ was additionally supported by the G\"{o}ran Gustafsson Foundation for Research in Natural Sciences and Medicine. 
C.D.L.\ is supported by the Science and Technology Facilities Council (STFC) [grant No. UKRI1172]. 
The work of Y.Z.\ is supported by NOIRLab, which is managed by the Association of Universities for Research in Astronomy (AURA) under a cooperative agreement with the U.S. National Science Foundation. 
J.d.V., I.S.-N., and L.T.S.C.\ are partially supported by the Spanish MICINN under grant PID2021-123012 and for the MAD4SPACE-CM TEC-2024/TEC-182 project funded by Comunidad de Madrid.
S.S.\ has received funding from the European Union’s Horizon 2022 research and innovation programme under the Marie Skłodowska-Curie grant agreement No 101105167 — FASTIDIoUS.
R.T.\ acknowledges co-funding from Next Generation EU, in the context of the National Recovery and Resilience Plan, Investment PE1 – Project FAIR ``Future Artificial Intelligence Research'', as well as Fondazione ICSC, Spoke 3 ``Astrophysics and Cosmos Observations'', Project ID CN00000013 ``Italian Research Center on High-Performance Computing, Big Data and Quantum Computing'', and partially supported by INFN INDARK grant. 
M.I.\ acknowledges support in part by the U.S. National Science Foundation under grant AST2327245, and also in part by the Department of Energy, office of Science, under Award Number DE-SC0022184. 
B.R.\ gratefully acknowledges the support of the NSF-Simons AI-Institute for the Sky (SkAI) via grants NSF AST-2421845 and Simons Foundation MPS-AI-00010513.
C.G.\ is funded by the MICINN project PID2022-141079NB-C32. IFAE is partially funded by the CERCA program of the Generalitat de Catalunya.
T.Z.\ is supported by Schmidt Sciences. 
J.Z.\ work is partially supported by Schmidt Futures, a philanthropic initiative founded by Eric and Wendy Schmidt as part of the Virtual Institute for Astrophysics (VIA).  
C.A. acknowledges support from DOE grant DE-SC009193.
A.G.\ is supported by an LSST-DA Catalyst Fellowship, made possible through the support of Grant 62192 from the John Templeton Foundation to LSST-DA. The opinions expressed in this publication are those of the author(s) and do not necessarily reflect the views of LSST-DA or the John Templeton Foundation.
T.T.\ acknowledges funding from the Swiss National Science Foundation under the Ambizione project PZ00P2\_193352. 
M.M.\ acknowledges support by the SNSF through return CH grant P5R5PT\_225598 and Ambizione grant PZ00P2\_223738. 
The work of A.A.P.M.\ was supported by the U.S. Department of Energy under contract number DE-AC02-76SF00515.
M.W.C.\ acknowledges support from the National Science Foundation with grant numbers PHY-2117997, PHY-2308862 and PHY-2409481. A.\'C.\, A.D.-W.\: This work was produced by FermiForward Discovery Group, LLC under Contract No.\ 89243024CSC000002 with the U.S. Department of Energy, Office of Science, Office of High Energy Physics. Publisher acknowledges the U.S.\ Government license to provide public access under the (\href{http://energy.gov/downloads/doe-public-access-plan}{DOE Public Access Plan}).
G.M.\ acknowledges support from Illinois Campus Research Board Award RB25035, NSF grant AST-2308174, and NASA grants 80NSSC24K0219 and 80NSSC25K7739. This work used Delta and DeltaAI at NCSA through allocations PHY240290, PHY250333, PHY250374, PHY250386, PHY250281 and PHY250308 from the Advanced Cyberinfrastructure Coordination Ecosystem: Services \& Support (ACCESS) program, which is supported by U.S. National Science Foundation grants \#2138259, \#2138286, \#2138307, \#2137603, and \#2138296. This work utilizes resources supported by the National Science Foundation's Major Research Instrumentation program, grant \#1725729, as well as the University of Illinois at Urbana-Champaign.
M.G.\ is supported by the European Union’s Horizon 2020 research and innovation programme under ERC Grant Agreement No.\ 101002652. 
Argonne National Laboratory’s work was supported under the US Department of Energy contract DE-AC02-06CH11357.
T.S.\ gratefully acknowledge the support of the NSF-Simons AI-Institute for the Sky (SkAI) via grants NSF AST-2421845 and Simons Foundation MPS-AI-00010513. T.S.\ was supported by NSF through grant AST-2510183 and by NASA through grants 22-ROMAN22-0055 and 22-ROMAN22-0013.
Y.-Y.M. is in part supported by the U.S. Department of Energy, Office of Science, Office of High Energy Physics, Experimental Research at the Cosmic Frontier program under Award Number DE-SC0009959.
% Individual acknowledgments
We are thankful to DESC members Pierre-Francois Leget, Prakruth Adari, Mariano Dominguez, and Xiangchong Li for their contributions and feedback in the preparation of this manuscript.